Classless Java

### 2016-03-24 OOPSLA16/sections/Introduction.tex
Introduction

Object oriented languages strive to offer great code reuse.
They aim to couple flexibility and rigour, expressive power and
modular reasoning.  Two main ideas have emerged to this end: prototype
based (PB)~\cite{Ungar87self} and class based (CB) languages such as
Java or Scala.  In prototype based
languages objects inherit from other objects, and thus objects own
both behaviour and state (and objects are all you have).
In class based languages an object is an instance of a specific class,
and classes inherit from other classes.  Here objects own the state,
while classes contain behaviour and the structure of the state.

@@

One important advantage of the use of abstract state operations and
automatic type-refinement is that it offers a new form of
\emph{type-safe covariant mutable state}. That is, in IB programming,
it is possible to refine \emph{mutable} ``fields'' in subtypes. This is
typically forbidden in CB programming languages (including Java and
Scala), since it is widely known that \emph{naive} type-refinement of
mutable fields is not type-safe. While covariant refinement of mutable
fields is supported by some type systems~\cite{bruce98astatically,bruce1994paradigmatic,ernst06virtual,Saito2013933}, this requires
significant complexity and restrictions to ensure that all uses of
covariant state are indeed type-safe.

@@

Using Java annotation processors, we produce an implementation of
Classless Java, which allow us to stick to pure Java 8. By annotating
with \Q|@Obj| the interfaces that represent object interfaces, we can
generate boring and repetitive code for interface instantiation and
type refinement. Such code should not be needed in the first place in
a real IB language, and the annotation processor allows us to
transparently hide it from Java programmers.  The implementation works
by performing AST rewriting, allowing most existing Java tools (such as
IDEs) to work out-of-the-box with our implementation. Moreover, the
implementation blends Java's conventional CB style and IB smoothly.
As a result, we experiment object interfaces with several interesting
Java programs, and conduct various case studies.  Finally, we also
formally define the behaviour of our \mixin annotation and discuss
its properties.

@@

To evaluate the usefulness of object interfaces, we illustrate
\numOfCaseStudies \bruno{needs updates}\yanlin{updated} 
applications. The first application is a simple 
solution to the Expression Problem~\cite{wadler98expression}, supporting independent 
extensibility~\cite{zenger05independentlyextensible}, and without boilerplate code. The second
application shows how embedded DSLs using fluent interfaces~\cite{fowler2005fluentinterface} 
can be easily defined using object interfaces. The third
application is a case study for a simple Maze game implemented with 
multiple inheritance. For this application we show that there is a
significant reduction in the number of lines of code when compared 
to an existing implementation~\cite{bono14} using plain Java 8. The last
application is a relatively larger case study of refactoring of a simple interpreter, showing that our
approach can benefit modularity and scale to real code base. Noteworthy, the first three applications are implemented 
without defining a single class! The last one contains several classes that cannot be 
converted to interfaces due to the limitation of our implementation of \mixin.

@@

\item {\bf Type-Safe Covariant Mutable State:} We show how the 
 combination of abstract state operations and type-refinement enables 
 a form of mutable fields that can be covariantly refined in a type-safe way.

@@

\item {\bf Case Studies:} We illustrate the usefulness of IB
  through various examples and case studies.

>>>>

Object oriented languages strive to offer great code reuse.
They aim to couple flexibility and rigour, expressive power and
modular reasoning.  Two main ideas emerged to this end: prototype
based (PB)~\cite{Ungar87self} and class based (CB) languages such as
Java or Scala.  In prototype based
languages objects inherit from other objects. Thus objects own
both behaviour and state (and objects are all you have).
In class based languages an object is an instance of a specific class,
and classes inherit from other classes.  Here objects own the state,
while classes contain behaviour and the structure of the state.

@@

One important advantage of the use of abstract state operations and
type-refinement is that it allows a new approach to
\emph{type-safe covariant mutable state}. That is, in IB programming,
it is possible to type-refine \emph{mutable} ``fields'' in subtypes. This is
typically forbidden in CB programming languages (including Java and
Scala), since it is widely known that \emph{naive} type-refinement of
mutable fields is not type-safe. While covariant refinement of mutable
fields is supported by some type systems~\cite{bruce98astatically,bruce1994paradigmatic,ernst06virtual,Saito2013933}, this requires
significant complexity and restrictions to ensure that all uses of
covariant state are indeed type-safe.

@@

Using Java annotation processors, we produce an implementation of
Classless Java, which allow us to stick to pure Java 8. By annotating
with \Q|@Obj| the interfaces that represent object interfaces, we can
automatically generate code for interface instantiation and
type refinement. Such code should not be needed in the first place in
a real IB language, and the annotation processor allows us to
transparently hide it from Java programmers.  The implementation works
by performing AST rewriting, allowing most existing Java tools (such as
IDEs) to work out-of-the-box with our implementation. Moreover, the
implementation blends Java's conventional CB style and IB smoothly.
As a result, we experiment object interfaces with several interesting
Java programs, and conduct various case studies.  Finally, we also
formally define the behaviour of our \mixin annotation and discuss
its properties.

@@

\begin{comment}
To evaluate the usefulness of object interfaces, we illustrate
\numOfCaseStudies \bruno{needs updates}\yanlin{updated} 
applications. The first application is a simple 
solution to the Expression Problem~\cite{wadler98expression}, supporting independent 
extensibility~\cite{zenger05independentlyextensible}, and without boilerplate code. The second
application shows how embedded DSLs using fluent interfaces~\cite{fowler2005fluentinterface} 
can be easily defined using object interfaces. The third
application is a case study for a simple Maze game implemented with 
multiple inheritance. For this application we show that there is a
significant reduction in the number of lines of code when compared 
to an existing implementation~\cite{bono14} using plain Java 8. The last
application is a relatively larger case study of refactoring of a simple interpreter, showing that our
approach can benefit modularity and scale to real code base. Noteworthy, the first three applications are implemented 
without defining a single class! The last one contains several classes that cannot be 
converted to interfaces due to the limitation of our implementation of \mixin.
\end{comment}

@@

\item {\bf Type-Safe Covariant Mutable State:} We show how the 
 combination of abstract state operations and type-refinement enables 
 a form of mutable state that can be covariantly refined in a type-safe way.

@@

\item {\bf Applications and Case Studies:} We illustrate the usefulness of IB
  through various examples and case studies.
